---
title: "DATA1901 Project 1"
author: "Aitchison 490378115, Braham 490422474, Hempstead 490424010"
subtitle: "The Influence of US College Majors"
date: "University of Sydney | Semester 1 2019 | Lab Session Monday 9am | Carslaw LS 353"
output: 
  html_document:
      code_folding: hide
      css: #https://use.fontawesome.com/releases/v5.0.6/css/all.css
      df_print: paged
      fig_caption: yes
      number_sections: yes
      self_contained: yes
      theme: journal #https://bootswatch.com
      toc: yes
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(ggplot2)) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require(plotly)) {
  install.packages("plotly")
  library(plotly)
}
```

<!--200 words Executive Summary, 400 words for the IDA, and 250 words (max) for each of the Research Questions.-->

# Executive Summary

<!-- need to cut down words --> 
*Education. Work. Money*: three words almost guaranteed to be at the forefront of a student's mind as they contemplate their future upon leaving highschool. With more and more opportunities for anyone to pursue almost any degree in any field, the inevitable question of *"Which major should I take?"* is becoming a harder and harder choice for students across the world. <br><br>
In this research project, data from over 6.7 million college graduates in the USA has been analysed to examine key questions regarding:

* Which college major should a student take to receive the highest income?
* Which college major should a student pursue to receive the greatest prospects of employment?
* How do these "rankings" align with the popularity of these courses?

Data shows that Engineering majors often see the highest levels of income, followed by other major categories such as Business and Law. In particular, Petroleum Engineering stands over the rest of the majors with a median income of USD\$110,000, compared to a median of \$36,000. <br> <br>

Further, looking at unemployment shows a trend in ... <br><br>

Ultimately, however, the popularity of courses appears not to be inherently linked to their employability prospects or expected incomes. This implies that, whether for better or worse, instead of basing choices on these factors, most students may instead be following influences such as personal preference, external pressure, or availabilities. 

<br>
<br>

# Full Report

## Initial Data Analysis (IDA)

```{r loadData, include=FALSE}
gradData = read.csv("data/recent-grads.csv")
```

### Domain Knowledge

1. In 2012, the median personal income for the US was [\$28,213](#medi), with unemployment at [8.1% for 2012](#unemp). 

2. Recently, a research article has shown that "since the [GFC]... students have turned away from the humanities and towards job-oriented degrees" [(Kopf, 2018)](#kopf), with the share of degrees in history dropping from 2% 2007 to 1% 2017 [(Kopf, 2018)](#kopf). This seems to reflect "a new set of student priorities... formed even before they see the inside of a college classroom... Students [are] fleeing humanities and related fields specifically because they think they have poor job prospects." [(Schmidt, 2018)](#schmidt).<br><br> 

### Source

The data was collected from the American Community Survey 2010 - 2012 Public Use Microdata Sample Files (PUMS) at the [USA Census Website](https://www.census.gov/programs-surveys/acs/data/pums.html). It was initially wrangled by media company [FiveThirtyEight](#fivethirty) (a part of ABC News Internet Ventures), with code accessible [here](https://github.com/fivethirtyeight/data/blob/master/college-majors/college-majors-rscript.R). <br><br>
The US Bureau of the Census is a government body, and although FiveThirtyEight had commercial interests, their process of data wrangling was highly transparent and reproducible. Therefore these sources can be considered reliable.

### Stakeholders
[The Census Bureau](#census) produces the PUMS as an inexpensive and accessible datasource for students and social scientists, while FiveThirtyEight wrangled this data for commercial use in their article [The Economic Guide to Picking a College Major](#economicguide), aimed at educating students on how to choose their college majors. <br> <br>
Drawing upon the domain knowledge, this data is particularly relevant to highschool leavers trying to choose a major, as well as current students contemplating their career prospects, as it may help them make a more informed economic decision. <br><br>
University staff, intership firms, or other organisations may also find benefit in predicting the future direction of the workforce, allowing for better resource allocation, such as investment into engineering and STEM fields. 

### Data Dictionary

```{r}
str(gradData)
```

This data consists of 20 variables (excluding "Rank" which orders the subjects by Median income), however, only 15 variables are relevant for the study:

#### Data Classifications {-.tabset}

##### Major {-}
The major's name.
<br>
*Type*: Factor 
<br>
*Assessment*: Either a character or factor classification would be suitable. 

##### Total {-}
Amount of total people with that major in the sample for 2010-2012.
<br>
*Type*: Integer 
<br>
*Assessment*: Suitable. 

##### Major_category {-}
General category for that major (e.g. "Engineering").
<br>
*Type*: Factor 
<br>
*Assessment*: Suitable - allows for easy classification and plotting.

##### Employed, Full_time, Part_time {-}
Number of people employed, employed 35 hours or more per week, and employed 35 hours or less respectively.
<br>
*Type*: Integer 
<br>
*Assessment*: Suitable. 

##### Full_time_year_round {-}
Number of people employed for at least 50 weeks per year and over 35 hours hours per week.
<br>
*Type*: Integer 
<br>
*Assessment*: Suitable. 

##### Unemployed {-}
Number of people considered unemployed by census data.
<br>
*Type*: Integer 
<br>
*Assessment*: Suitable. 

##### Unemployment_rate {-}
The percentage of people unemployed over (unemployed + employed).
<br>
*Type*: Number 
<br>
*Assessment*: Suitable. 

##### Median, P25th, P75th {-}
Median, 25th percentile, and 75th percentile earnings respectively for full-time, year-round workers (in USD).
<br>
*Type*: Integer 
<br>
*Assessment*: Suitable - although income is continuous, it can be considered discrete without significantly impacting the data. 

##### College_jobs, Non_college_jobs, Low_wage_jobs {-}
Number of people with a job requiring a college degree, not requiring a college degree, and in a low-wage service job respectively.
<br>
*Type*: Integer 
<br>
*Assessment*: Suitable. 



### Data Assessment

**Possible Issues**: <br>

 * The data consists of pre-summarised information, where each subject is a set of medians, percentages, etc. Therefore, care needs to be made in plotting and drawing conclusions. 
 * The data is categorised as an *observational study*, lacking in a control group and randomised allocation between majors. Therefore, any conclusions drawn must be treated as associativity rather than causality. 
 * The data is 7 years old, and income has not been adjusted for inflation. However, this is not a major issue considering relative comparisons are being made. 
 * The data has been pre-filtered to only include subjects below the age of 28. However, this can be positive in that it is more relevant to current university students.
 
<br>
**Validity**: <br>
This data, taking into account the issues above and their solutions, can be considered valid. However, care must be taken to acknowledge confounders, such as personality and circumstance, rather than just major choice, in influencing the variables. 

## Research Question 1

> Which college major should a student take to receive the highest income?

There are three variables to consider - the 25th percentile, median, and 75th percentile incomes. Additionally, it is important to consider both individual majors and major categories. Taking a summary initially shows that there is a significant range of incomes:
```{r}
# Shows the summary (IQR etc.) of median incomes
summary(gradData$Median)
```
```{r warning=FALSE}
# Used to plot the median line (in red)
hline <- function(y = 0) {
  list(
    type = "line", 
    x0 = 0, 
    x1 = 1, 
    xref = "paper",
    y0 = y, 
    y1 = y, 
    line = list(color = "red", width=1)
  )
}
# Plots a boxplot of major categories by income
plot_ly(gradData, y=~Median/1000, color=~Major_category, type="box") %>% 
  layout(    
    yaxis = list(title = "Median income (USD$1000)"),
    xaxis = list(showticklabels = FALSE),
    title = "Median Income per Major Category",
    shapes = list(hline(36)))
```
Plotting the median income against major category backs up the summary - showing a large spread, centred around the median of $36,000. 
<br>

```{r warning=FALSE, include=FALSE}
# Selects top 10
gradData.head = head(gradData, n=10) 
# Creates a new data frame to be able to plot median, 25th, and 75th percentiles on the same graph 
median.df = data.frame(Major = gradData.head$Major, Major_category=gradData.head$Major_category, 
                       Income = gradData.head$Median)
p25.df = data.frame(Major = gradData.head$Major, Major_category=gradData.head$Major_category, 
                       Income = gradData.head$P25th)
p75.df = data.frame(Major = gradData.head$Major, Major_category=gradData.head$Major_category, 
                       Income = gradData.head$P75th)
gradData.head.df = rbind(median.df, p25.df, p75.df)
# For reordering the majors
score = rbind(gradData.head, gradData.head, gradData.head)
# Plots a boxplot
plot_ly(gradData.head.df, y=~Income/1000, color=~reorder(sub("AND", "\nAND",Major), -score$Median), type="box") %>%
  layout(    
    yaxis = list(
      title = "Median income (USD$1000)",
      autotick = FALSE,
      ticks = "outside",
      tick0 = 0,
      dtick = 10,
      ticklen = 3),
    xaxis = list(showticklabels = FALSE),
    title = "Top 10 Majors by Income")
```

```{r warning=FALSE, include=FALSE}
# Selects bottom 10
gradData.tail = tail(gradData, n=10)
# Creates a new data frame to be able to plot median, 25th, and 75th percentiles on the same graph
median.df = data.frame(Major = gradData.tail$Major, Major_category=gradData.tail$Major_category, 
                       Income = gradData.tail$Median)
p25.df = data.frame(Major = gradData.tail$Major, Major_category=gradData.tail$Major_category, 
                       Income = gradData.tail$P25th)
p75.df = data.frame(Major = gradData.tail$Major, Major_category=gradData.tail$Major_category, 
                       Income = gradData.tail$P75th)
gradData.tail.df = rbind(median.df, p25.df, p75.df)
# For reordering the majors
score = rbind(gradData.tail, gradData.tail, gradData.tail)
# Plots a boxplot
plot_ly(gradData.tail.df, y=~Income/1000, color=~reorder(sub("DISORDERS SCIENCES \nAND", "DISORDERS\nSCIENCES AND", sub("AND", "\nAND",Major)), -score$Median), type="box") %>%
  layout(    
    yaxis = list(
      title = "Median income (USD$1000)",
      autotick = FALSE,
      ticks = "outside",
      tick0 = 0,
      dtick = 2.5,
      ticklen = 3,
      tickwidth = 1),
    xaxis = list(showticklabels = FALSE),
    title = "Bottom 10 Majors by Income")

```

```{r fig.height=8, warning=FALSE}
# Selects top 10
gradData.head = head(gradData, n=10)
# Creates a new data frame to be able to plot median, 25th, and 75th percentiles on the same graph
median.df = data.frame(Major = gradData.head$Major, Major_category=gradData.head$Major_category, 
                       Income = gradData.head$Median)
p25.df = data.frame(Major = gradData.head$Major, Major_category=gradData.head$Major_category, 
                       Income = gradData.head$P25th)
p75.df = data.frame(Major = gradData.head$Major, Major_category=gradData.head$Major_category, 
                       Income = gradData.head$P75th)
gradData.head.df = rbind(median.df, p25.df, p75.df)
# Selects bottom 10
gradData.tail = tail(gradData, n=10)
# Creates a new data frame to be able to plot median, 25th, and 75th percentiles on the same graph
median.df = data.frame(Major = gradData.tail$Major, Major_category=gradData.tail$Major_category, 
                       Income = gradData.tail$Median)
p25.df = data.frame(Major = gradData.tail$Major, Major_category=gradData.tail$Major_category, 
                       Income = gradData.tail$P25th)
p75.df = data.frame(Major = gradData.tail$Major, Major_category=gradData.tail$Major_category, 
                       Income = gradData.tail$P75th)
gradData.tail.df = rbind(median.df, p25.df, p75.df)
score = rbind(gradData.tail, gradData.tail, gradData.tail)
# Combines the two 
gradData.combined.df =rbind(gradData.head.df, gradData.tail.df)
# Orders it
score = rbind(gradData.head, gradData.head, gradData.head, score)
# Plots a boxplot
plot_ly(gradData.combined.df, y=~Income/1000, x=~reorder(Major, -score$Median), color=~Major_category, type="box") %>%
  layout(    
    yaxis = list(
      title = "Median income (USD$1000)",
      autotick = FALSE,
      ticks = "outside",
      tick0 = 0,
      dtick = 10,
      ticklen = 3,
      tickwidth = 1,
      tickwidth = 1),
    xaxis = list(
      showticklabels = TRUE, title="",
      tickangle = 270, tickfont = list(size = 10)),
    title = "Top 10 and Bottom 10 Majors by Median Income")
```
Looking at individual majors, there are initially too many data-points to make sense of the information. Instead, ordering the data by median income, the subjects can be limited to only the top and bottom 10 majors (note that this plotted data takes into account median, 25th percentile, and 75th percentile). While 9 of the top 10 majors belong to the Engineering category, the bottom 10 majors are considerably more varied. 

<br>
```{r fig.height=6, fig.width=8}
# Selects all of gradData
combined = gradData 
# Combines median, 25th percentile, and 75th percentile into one data frame for plotting
median.df = data.frame(Major = combined$Major, Major_category=strtrim(combined$Major_category, 45), 
                       Income = combined$Median)
p25.df = data.frame(Major = combined$Major, Major_category=strtrim(combined$Major_category, 45), 
                       Income = combined$P25th)
p75.df = data.frame(Major = combined$Major, Major_category=strtrim(combined$Major_category, 45), 
                       Income = combined$P75th)
combined.df = rbind(median.df, p25.df, p75.df)
# Plots a density chart
ggplotly(ggplot(combined.df, aes(x=Income/1000, fill=Major_category)) + geom_density(alpha=0.2) +
           facet_wrap(~Major_category) +
           xlab("Income (USD$1000)") + ylab("Density") + 
           labs(title="Income Distribution per Major Category") + theme_minimal() + 
           theme(legend.position="none", strip.text.x = element_text(size = 7), 
                 axis.text.y = element_blank()))
```
```{r}
# Selects only Engineering, education and business majors
combined = rbind(gradData[gradData$Major_category=="Engineering",],
                 gradData[gradData$Major_category=="Education",],
                 gradData[gradData$Major_category=="Business",])
# Combines median, 25th percentile, and 75th percentile into one data frame for plotting
median.df = data.frame(Major = combined$Major, Major_category=combined$Major_category, 
                       Income = combined$Median)
p25.df = data.frame(Major = combined$Major, Major_category=combined$Major_category, 
                       Income = combined$P25th)
p75.df = data.frame(Major = combined$Major, Major_category=combined$Major_category, 
                       Income = combined$P75th)
combined.df = rbind(median.df, p25.df, p75.df)
# Plots a density chart
ggplotly(ggplot(combined.df, aes(x=Income/1000, fill=Major_category)) + geom_density(alpha=0.2) +
           xlab("Income (USD$1000)") + ylab("Density") + 
           labs(fill="Major Category", title="Income Distribution per Major Category (Selection)") +
           theme_minimal())
```

Examining the density estimation of a selection of major categories, again Engineering appears to have significantly higher incomes compared to other categories. However, the graph shows that the spread is also significantly larger, with a portion of the income falling within the range of the lowest majors. This is contrasted with Education, where the range is confined to ~$25,000.  
```{r}
# Coefficient of Variation for the Engineering sample's incomes
message("Coefficient of Variation for Engineering: ", sd(combined.df[combined.df$Major_category=="Engineering",]$Income)/
  mean(combined.df[combined.df$Major_category=="Engineering",]$Income)) 
# Coefficient of Variation for the Education sample's incomes
message("Coefficient of Variation for Education: ", sd(combined.df[combined.df$Major_category=="Education",]$Income)/
  mean(combined.df[combined.df$Major_category=="Education",]$Income)) 
```
This is re-iterated by the coefficient of variation for Engineering being over 150% of Education's. 
<br> <br>
**Summary**: <br>
The data shows that Engineering incomes can far exceed those in other categories, with Petroleum Engineering in particular being significantly higher than the other majors. Indeed, the separation of Petroleum Engineering from the other top 10 median incomes is comparable to the separation of the top 10 from the bottom 10. However, Engineering incomes overall have a significantly larger spread than the other categories, implying a volatility either between majors or within the industries themselves. Nevertheless, students seeking high incomes may be best suited to look towards Engineering fields. 
<br><br>

## Research Question 2
> Which college major should a student pursue to see the greatest prospects for employment? 
> Which college majors prove most useful in achieving employment?

Look at variables related to employment (both employment chances, and the need for a college degree in relation to employment)

Compare unemployment rates between majors and major categories
Compare worth of college degree in achieving employment

The fact that this data is for <28 changes things lots. It's bascially entry level income and does not reflect increasing income over time.


<br>
Insert text and analysis. 

```{r warning=FALSE}

```


```{r}
percentCollege = gradData$College_jobs/(gradData$College_jobs+gradData$Non_college_jobs+gradData$Low_wage_jobs)
ggplot(gradData, aes(x=factor(Major_category), y=percentCollege)) + 
  geom_boxplot(aes(fill = factor(Major_category))) +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(fill = "Major Category", title="Percent College Jobs per Major Category") + 
  ylab("Percentage of Jobs as College Jobs")
```
<br>How much of this employment is actually in their field / based on the degree (looking at college jobs vs non-college jobs)?

<br>

## Research Question 3

> Looking at the results from Question 1 and Question 2, how do these "rankings" align with the popularity of these courses? 

From Questions 1 and 2, the majors falling into the engineering category have consistently ranked among the highest in terms of income. By examining the popularity of courses, a stronger conclusion should be able to be made regarding the reasons behind this.

```{r message=FALSE}
# Exclude Food Science and Military Technologies, as they have missing data.
# total_grads = sum(gradTotalData$Total)
# Order our data by the total number of graduates
gradTotalData = gradData[!(gradData$Major=="FOOD SCIENCE") & !(gradData$Major=="MILITARY TECHNOLOGIES") ,]

gradTotalData = gradTotalData[order(gradTotalData$Total),]

bottom_ten_by_total = head(gradTotalData, 10)
top_ten_by_total = tail(gradTotalData, 10)

gradTotal_topbottom = rbind(top_ten_by_total, bottom_ten_by_total)


ggplotly(ggplot(gradTotalData, aes(x=Median/1000, fill=Major_category)) + geom_histogram() + 
           theme(legend.position = "none") +
           xlab("Median Income (USD$1000)") + ylab("Count") + 
           labs(title="Popularity vs Median Income"))
```
Examining major categories (differentiated by colour, and named on hover), the data shows that engineering has a higher frequency in higher median income areas. Despite this, the data still follows a relatively normal distribution.
```{r message=FALSE}
ggplotly(ggplot(gradTotalData, aes(x=100*Employed/Total, fill=Major_category)) + geom_histogram() + 
           theme(legend.position = "none") +
           xlab("Employment Rate (%)") + ylab("Count") + labs(title="Popularity vs Employment Rate"))
```
Engineering holds not only the highest median income major, but also the major with the highest employment rate of graduates. However, employment rates of engineering majors are also observed to have much more of a spread. As shown below, the employment rate of engineering majors at worst is 2.75 deviations from the mean.
```{r}
employmentMean = mean((gradTotalData$Employed/gradTotalData$Total))
employmentSd = sd((gradTotalData$Employed/gradTotalData$Total))
# Get all employment rates of engineering, then sort them in increasing order, and take the first element to get the lowest.
lowestEmployment = sort((gradTotalData$Employed/gradTotalData$Total)[gradTotalData$Major_category=="Engineering"], decreasing=FALSE)[1]
distance = ((employmentMean - lowestEmployment)/employmentSd)
message("Lowest Z Score for Engineering Employment: ", distance)
```
```{r}
ggplotly(ggplot(gradTotal_topbottom, aes(x=Median/1000, fill=Major, y=Total)) + geom_point() + 
           theme(legend.position = "none") +
           xlab("Median Income (USD $1000)") + ylab("Total Grads") + labs(title="Total Grads vs Median Income"))
```
By examining the top and bottom ten majors as a scatter plot of popularity against income, the normal distribution is of income is emphasised, suggesting that students may not be heavily influenced by financial prospects alone. <br><br>
**Summary**: <br>

From this data, it appears that popularity in the courses is not largely influenced by their employability prospects or expected incomes. Instead, both income and employment seem to follow a relatively normal distribution curve. This is in line with what would be expected if the major choices were randomly distributed. Thus, it can be concluded that the choices for majors are affected by a large range of hidden factors. Further research would have to be done to investigate the impact of these factors.

<br>

# References

<a name="unemp"></a> 
Bureau of Labor Statistics. (2019). *Labor Force Statistics from the Current Population Survey, 2012 (LNU04000000)* [Data set]. Retrieved from [http://data.bls.gov](http://data.bls.gov). <br> <br>
<a name="economicguide"></a>
Casselman, B. (2014, September 12). The Economic Guide to Picking a College Major. *FiveThirtyEight*. Retrieved from [https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/). <br><br>
<a name="fivethirty"></a>
FiveThirtyEight. (2014). *College Majors 2010-2012  (Recent Grads)* [Data set]. Retrieved from Github; [https://github.com/fivethirtyeight/data/tree/master/college-majors](https://github.com/fivethirtyeight/data/tree/master/college-majors). <br> <br>
<a name="kopf"></a>
Kopf, D. (2018, August 29). The 2008 financial crisis completely changed what majors students choose. *Quartz*. Retrieved from [https://qz.com/1370922/the-2008-financial-crisis-completely-changed-what-majors-students-choose/](https://qz.com/1370922/the-2008-financial-crisis-completely-changed-what-majors-students-choose/). <br><br>
<a name="schmidt"></a>
Schmidt, B. (2018, August 3). The Humanities Are in Crisis. *The Atlantic*. Retrieved from [https://www.theatlantic.com/ideas/archive/2018/08/the-humanities-face-a-crisisof-confidence/567565/](https://www.theatlantic.com/ideas/archive/2018/08/the-humanities-face-a-crisisof-confidence/567565/). <br><br>
<a name="census"></a>
US. Bureau of the Census. (2018). *Public Use Microdata Samples (PUMS) Documentation*. Retrieved from [https://www.census.gov/programs-surveys/acs/technical-documentation/pums.html](https://www.census.gov/programs-surveys/acs/technical-documentation/pums.html). <br><br>
<a name="medi"></a> 
U.S. Bureau of the Census. (2017). *Real Median Personal Income in the United States, 2012 (MEPAINUSA672N)* [Data set]. Retrieved from FRED, Federal Reserve Bank of St. Louis; [https://fred.stlouisfed.org/series/MEPAINUSA672N](https://fred.stlouisfed.org/series/MEPAINUSA672N). <br><br>


```{r}
sessionInfo()
```


