---
title: "DATA1901 Project 1"
author: "Aitchison 490378115, Braham 490422474, Hempstead XXXXXXXXX"
subtitle: "On the Influence of US College Majors"
date: "University of Sydney | Semester 1 | March 2019"
output: 
  html_document:
      code_folding: show
      css: #https://use.fontawesome.com/releases/v5.0.6/css/all.css
      df_print: paged
      fig_caption: yes
      number_sections: yes
      self_contained: yes
      theme: yeti
      toc: yes
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--We suggest 400 words (max) for the Executive Summary, 200 words (max) for the IDA, and 250 words (max) for each of the Research Questions.-->

# Executive Summary

<br>
<br>

# Full Report

## Initial Data Analysis (IDA)

```{r}
gradData = read.csv("data/recent-grads.csv")
```


### Source
<!--Hmmmm--> 
```{r, include=FALSE}
library(plotly)
```

```{r}
p2 = plot_ly(mtcars, x = ~mpg, y = ~wt, z = ~qsec, color = ~factor(cyl), type="scatter3d", mode="markers", text = ~paste("Cylinder: ", cyl)) %>%
  layout(
    title = "3D Chart for mtcars",
    scene = list(
      xaxis = list(title = "Miles per gallon"),
      yaxis = list(title = "Weight"),
      zaxis = list(title = "Qsec")
    ))
p2
```


### Data Dictionary

### Data Assessment

- The data is valid because ...
- Possible issues include ...
- We will solve this by ...

<br>

## Research Question 1

Insert text and analysis. 

Summary: 

## Research Question 2

Insert text and analysis. 

Summary: 

## Research Question 3

Insert text and analysis. 

Summary: 

<br>

# Conclusions

Insert text.

<br>

# References

Style: APA 

<br>

# Personal reflection on group work

- The way I contributed was ...
- What I learnt about group work was ... 

<br>

