---
title: "DATA1901 Project 1"
author: "Aitchison 490378115, Braham 490422474, Hempstead XXXXXXXXX"
subtitle: "On the Influence of US College Majors"
date: "University of Sydney | Semester 1 | March 2019"
output: 
  html_document:
      code_folding: show
      css: #https://use.fontawesome.com/releases/v5.0.6/css/all.css
      df_print: paged
      fig_caption: yes
      number_sections: yes
      self_contained: yes
      theme: journal #https://bootswatch.com
      toc: yes
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require(ggplot2)) {
  install.packages(ggplot2)
  library(ggplot2)
}
```

<!--We suggest 400 words (max) for the Executive Summary, 200 words (max) for the IDA, and 250 words (max) for each of the Research Questions.-->

# Executive Summary

<br>
<br>

# Full Report

## Initial Data Analysis (IDA)

```{r loadData}
gradData = read.csv("data/recent-grads.csv")
```


### Source

Data collected from the American Community Survey 2010 - 2012 Public Use Microdata Series (PUMS Data) at the [USA Census Website](https://www.census.gov/programs-surveys/acs/data/pums.html). It was initially wrangled by American media company [FiveThirtyEight](https://github.com/fivethirtyeight/data/blob/master/college-majors/college-majors-rscript.R). 

Major categories were sourced from ["What's It Worth?: The Economic Value of College Majors." Georgetown University Center on Education and the Workforce, 2011](http://cew.georgetown.edu/whatsitworth). 

Note that:

- Data was not adjusted for inflation 
- "Recent-grads" data was filtered to only include subjects below the age of *28*

### Data Dictionary

<!--TODO: Make this table wider, retype in own words, add classifications of the data--> 
Header | Description
---|-------
`Rank` | Rank by median earnings
`Major_code` | Major code, FO1DP in ACS PUMS
`Major` | Major description
`Major_category` | Category of major from Carnevale et al
`Total` | Total number of people with major
`Sample_size` | Sample size (unweighted) of full-time, year-round ONLY (used for earnings)
`Men` | Male graduates
`Women` | Female graduates
`ShareWomen` | Women as share of total
`Employed` | Number employed (ESR == 1 or 2)
`Full_time` | Employed 35 hours or more
`Part_time` | Employed less than 35 hours
`Full_time_year_round` | Employed at least 50 weeks (WKW == 1) and at least 35 hours (WKHP >= 35)
`Unemployed` | Number unemployed (ESR == 3)
`Unemployment_rate` | Unemployed / (Unemployed + Employed)
`Median` | Median earnings of full-time, year-round workers
`P25th` | 25th percentile of earnigns
`P75th` | 75th percentile of earnings
`College_jobs` | Number with job requiring a college degree
`Non_college_jobs` | Number with job not requiring a college degree
`Low_wage_jobs` | Number in low-wage service jobs

### Data Assessment

- The data is valid because ...
- Possible issues include ...
- We will solve this by ...

<br>

## Research Question Drafts Old

Q1
"Is there a correlation between majors that promote higher paying jobs and the associated unemployment rates?" 
Scatter plot (LOB)
(Doesn't really work :(((( )

Q1B
"Is there a correlation between gender ratios and expected income?"

Q2
"Which college degrees are most relevant to the subsequent job (based on number of college jobs)?" 

Q3 *Based on previous results
"What are the comparative unemployment rates between Engineering and Liberal Arts graduates?" 
Unemployment rate between major categories (Simpson's Paradox) potential
Get top 10 from each Humanities and Liberal Arts + Engineering 

# #New questions:
Q1: Which college majors give the highest incomes?
Sub-question: How does this take into account spread (i.e. compare 75th vs median vs 25th)

Q2: Which college majors give the highest employability (based on unemployment)?
Sub-question: How much of this employment is actually in their field / based on the degree (looking at college jobs vs non-college jobs)?

Q3: Which college majors pay the most per time worked (looking at % full time)? 

Q3 (alternative): Looking at the results from Q1 and Q2, how do these "rankings" align with the popularity of these courses (based on total people = employed + unemployed)

## Research Question 1

Insert text and analysis. 

Summary: 

```{r}

```

## Research Question 2

Insert text and analysis. 

Summary: 

```{r}
percentCollege = gradData$College_jobs/(gradData$College_jobs+gradData$Non_college_jobs+gradData$Low_wage_jobs)
ggplot(gradData, aes(x=factor(Major_category), y=percentCollege)) + 
  geom_boxplot(aes(fill = factor(Major_category))) +
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(fill = "Major Category", title="Percent College Jobs per Major Category") + 
  ylab("Percentage of Jobs as College Jobs")
```

## Research Question 3

Insert text and analysis. 

Summary: 

<br>

# Conclusions

Insert text.

<br>

```{r}
sessionInfo()
```


# References

Style: APA 

<br>

# Personal reflection on group work

- The way I contributed was ...
- What I learnt about group work was ... 

<br>

